---
title: "Hello Census"
author: "maevadevs"
date: "12/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import and Setups

Load the tidycensus package into your R session

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(tidycensus)
```

Load all our helper functions: 
- `get_key` is a custom function that allows to load the keys from `.env`
- All env settings are stored in `.env` file

```{r echo=TRUE, message=FALSE, warning=FALSE}
source('../helper-functions.R')
envfile = "../.env"
```

Get the census api key from the file

```{r echo=TRUE, message=FALSE, warning=FALSE}
api_key = get_key(envfile, "census_api_key")
```

Set Census API Key with `census_api_key()`: This is part of tidycensus

```{r echo=TRUE, message=FALSE, warning=FALSE}
census_api_key(api_key)
```

We could check the API key value with the following

```r
print(Sys.getenv("CENSUS_API_KEY"))
```

---

## Functions from `tidycensus`

- `get_decennial()`: As default, return data from the 2010 US Census
- `get_acs()`: As default, return data from 2012-2016 ACS. Grants access to the 1-year and 5-year American Community Survey APIs
  - Access to the 1990 (SF1 and SF3) and 2000 (SF3) APIs has been suspended

Obtain and view state populations from the 2010 US Census

```{r echo=TRUE, message=FALSE, warning=FALSE}
state_pop <- get_decennial(geography = "state", variables = "P001001")
state_pop %>% head
```

Obtain and view state median household income from the 2012-2016 American Community Survey (ACS)

```{r echo=TRUE, message=FALSE, warning=FALSE}
state_income <- get_acs(geography = "state", variables = "B19013_001")
state_income %>% head
```

- Census data comprise thousands of variables available across dozens of geographies
  - Most are available through `tidycensus`
  - But it helps to understand the options
- Census Tract: Only available by state
  - `tidycensus` can help to convert between state and county
- `tidycensus` includes the Census variable ID in the variable column
  - If a user want to supply their own variable name, this can be accomplished with a named vector

Get an ACS dataset for Census tracts in Texas by setting the state

- `B19013_001`: Median household income

```{r echo=TRUE, message=FALSE, warning=FALSE}
tx_income <- get_acs(geography = "tract",
                     variables = "B19013_001",
                     state = "TX")
tx_income %>% head
```

Get an ACS dataset for Census tracts in Texas/Travis County by setting the state and county

```{r echo=TRUE, message=FALSE, warning=FALSE}
travis_income <- get_acs(geography="tract",
                     variables="B19013_001",
                     state="TX",
                     county="Travis")
travis_income %>% head
```

We can also set a better alias name for the variable using a vector

```{r echo=TRUE, message=FALSE, warning=FALSE}
travis_income <- get_acs(geography="tract",
                     variables=c(household_income="B19013_001"),
                     state="TX",
                     county="Travis")
travis_income %>% head
```

- By default, tidycensus functions return tidy data frames
- But we can change that to wide dataframe as well
- This is useful when we want to analyze multiple variables

```{r echo=TRUE, message=FALSE, warning=FALSE}
oregon_wide <- get_acs(geography="tract",
                       state="OR",
                       variables=c(hhincome = "B19013_001", medage="B01002_001"),
                       output = "wide")
oregon_wide %>% head
```

Create a scatterplot

```{r echo=TRUE, message=FALSE, warning=FALSE}
plot(oregon_wide$hhincomeE, oregon_wide$medageE)
```

---

## Searching for Data with `tidycensus`

- To use `tidycensus` functions, user must supply a vector of census IDs

### Searching for Census Variables

- [Census Reporter](https://censusreporter.org/)
- Built-in tidycensus variable searching: 

### `load_variables(year, dataset, cache?)`

- Return can be explored with `tidyverse`
- Return columns:
  - `name`: Census Variable ID
  - `label`: Variable description for human-readable
  - `concept`: General group to which the variable corresponds

Load variables

```{r echo=TRUE, message=FALSE, warning=FALSE}
v16 <- load_variables(year=2016,
                      dataset='acs5', # or acs5/profile, sf3...
                      cache=TRUE)
v16 %>% head
```

Filter with `tidyverse`

```{r echo=TRUE, message=FALSE, warning=FALSE}
B19001 <- v16 %>% filter(str_detect(name, "B19001"))
B19001 %>% head
```

We can also filter by a different column

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Use public transportation to search for related variables
public_trans <- v16 %>% filter(str_detect(label,
                                  fixed("public transportation",
                                        ignore_case = TRUE)))
public_trans %>% head
```

### Typical Variable Format

`[Table ID]_[Column Number]_[Row Number][Variable Type]`

- The variable type is optional for `tidycensus`
  - `tidycensus` returns both `E` (Estimate) and `M` (Margin of Error) by default
  - Suffixes are only showing up when returning in wide-format

---

## Visualizing Census Data With `ggplot2`

Median household income in Northeast from the 2016 ACS

```{r echo=TRUE, message=FALSE, warning=FALSE}
ne_income_16 <- get_acs(geography="state",
                        variables=c("B19013_001"),
                        survey="acs1",
                        state=c("ME","NH","VT","MA","RI","CT","NY"))
ne_income_16 %>% head
```

Now, we can plot the data using ggplot2

```{r echo=TRUE, message=FALSE, warning=FALSE}
ne_income_16 %>% 
  ggplot(aes(
    x=estimate,
    y=reorder(NAME, estimate) # Reorder y in the increasing of x
  )) + 
  geom_point(color="navy", size=4) +
  scale_x_continuous(labels=scales::dollar) +
  theme_minimal(base_size=14) +
  labs(x="2016 ACS Estimate",
       y="",
       title="Median Household Income By State")
```







